Sperm do not evolve to collaborate in female meiotic drive.

### 2014-04-30 writeup_and_notes/male_dependent_drive.tex
fixed minor typos, changes in text inequalities to single-line rather than fract formatting

In the standard single-locus, biallelic model of female meiotic drive, the driving allele is transmitted to the egg  in heterozygotes with probability  $d > \frac{1}{2}$, regardless of sperm genotype. 
To depict a case of a self-promoting meiotic
driver,  we modify this standard model such that the driver is only effective when fertilized by a sperm carrying that allele (see Figure
\ref{Eggsperm_2_allele_cartoon}). 
%\footnote{the timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
%	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
%	so $d$ is bounded to be $<0.75$? \citep[see ][ for discussion]{Buckler1999}. Could also put this in
  %      caption of fig.}.
We then identify the conditions allowing the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
%\yb{ We present a general framework for this problem in equations A1-A4, and evaluate specialized cases below.}

For comparative purposes, we present the standard drive model in equations \ref{eqn.A1}-\ref{eqn.A3}, 
	and refer readers to the previous literature on this model (\citep[e.g. ][]{Prout1973,Ubeda2004} for additional results). 
In this case, assuming the driving allele is deleterious but fully recessive 
	(i.e. homozygous fitness equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large drivers cannot fix \citep{Prout1973}, 
	providing an opportunity for the evolution of
	drive suppressors, corresponding well to 
	empirical examples of female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
	a self-promoting driver specifically creates homozygotes by driving. 
%, and therefore there is a drive-associated fitness cost even when rare. 
Thus it must overcome a drive-associated homozygous fitness cost simply to spread when rare,
	and the conditions allowing the invasion of a self-promoting driver
 	are far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than $\frac{2 d - 1}{4 d}$. 
This analytical result, derived  from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Invasion_space}).

@@

The positive frequency dependence of a self-promoting driver can induce bistability in its spread. 
When rare, a  recessive self-promoting driver can spread only if $s < \frac{2 d - 1}{4 d}$;
	however, once common such a driver can fix over a broader range of parameter space (i.e. if $s<\frac{2d-1}{2}$). 
Therefore, when $\frac{2 d - 1}{4 d}<s<\frac{2d-1}{2}$, 
	a recessive self-promoting driver will deterministically fix if drift, mutation, 
	or migration pressure bring its frequency above $\frac{1-2d+4ds}{2s(2d-1)}$, 
	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-h_s$)
	%(e.g. additive fitness cost) 
	further constrains the evolution of a self-promoting driver. 
This case also displays bistability -- a rare self-promoting driver cannot spread; 
	however, if $s < \frac{2 d - 1+s_h(3-2d)}{2}$ it will fix deterministically if its 
	frequency is greater than some threshold, $f_B^*$ [Equation \eqref{thresh2}, and \yb{FIGURES SXX}].
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that, if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be observed as polymorphisms in natural populations.

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm).
%GRAHAM I HAVE REWRITTEN THIS SECTION TO REFLECT MATHS AND MY PARAMETERIZATION. WE CAN WORK ON SYNCHING OUR NOTATION
%This adds an additional parameter to the system -- the dominance of the drive enhancer in heterozygous males, 
%	when this dominance is high, a narrow sliver of parameter space sustaining a polymorphism emerges (see Figure S\ref{Effect_of_dominance})
 %presumably because positive frequency dependence diminishes with the dominance of the male effect driver.
This male-genotype dependent model requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
When $d_{het}$ is large relative to $d_{hom}$, a narrow sliver of parameter space sustains a polymorphism (see Figure S\ref{Effect_of_dominance}), presumably because positive frequency dependence diminishes with the dominance of the male effect driver. 
As an analytical example, consider the case of male-genotype dependent deleterious recessive female driver 
	(assuming, as above, $w_{BB}=1-s$, and $w_{AA}=w_{AB}=1$). 
Here, a rare driver invades when $s<\frac{d_{het}-1/2}{d_{het}}$, and fixes when $s < d_{hom} - \frac{1}{2}$. 
Intermediate values of $s$, which represent a small proportion of parameter space can maintain a polymorphism.

>>>>

In the standard single-locus, biallelic model of female meiotic drive, the driving allele is transmitted to the egg  in heterozygotes with probability  $d > 1/2$, regardless of sperm genotype. 
To depict a case of a self-promoting meiotic
driver,  we modify this standard model such that the driver is only effective when fertilized by a sperm carrying that allele (see Figure
\ref{Eggsperm_2_allele_cartoon}). 
%\footnote{the timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
%	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
%	so $d$ is bounded to be $<0.75$? \citep[see ][ for discussion]{Buckler1999}. Could also put this in
  %      caption of fig.}.
We then identify the conditions allowing the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
%\yb{ We present a general framework for this problem in equations A1-A4, and evaluate specialized cases below.}

For comparative purposes, we present the standard drive model in equations \ref{eqn.A1}-\ref{eqn.A3}, 
	and refer readers to the previous literature on this model \citep[e.g. ][]{Prout1973,Ubeda2004} for additional results. 
In this case, assuming the driving allele is deleterious but fully recessive 
	(i.e. homozygous fitness equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large drivers cannot fix \citep{Prout1973}, 
	providing an opportunity for the evolution of
	drive suppressors, corresponding well to 
	empirical examples of female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
	a self-promoting driver specifically creates homozygotes by driving. 
%, and therefore there is a drive-associated fitness cost even when rare. 
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than $(2 d - 1)/(4 d)$. 
This analytical approximation, derived  from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Invasion_space}).

@@

The positive frequency dependence of a self-promoting driver can induce bistability in its spread. 
Compared to the invasion criteria, parameter space allowing fixations is much broader -- 
	a common, recessive, self-promoting driver fixes if  $s$ is less than $(2d-1)/2$. 
Therefore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, 
	a recessive, self-promoting driver will deterministically fix if drift, mutation, 
	or migration pressure bring its frequency above $(1-2d+4ds)/(2s(2d-1))$, 
	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-h_s$)
	%(e.g. additive fitness cost) 
	further constrains the evolution of a self-promoting driver. 
This case also displays bistability -- with any heterozygous fitness cost, a rare self-promoting driver cannot spread. 
However, if $s < d - 1/2+s_h(3-2d)/2$, 
	it will fix deterministically if its 
	frequency is greater than a threshold, $f_B^*$ [Equation \eqref{thresh2}, and \yb{FIGURES SXX}].
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that, if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be observed as polymorphisms in natural populations.

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
%GRAHAM I HAVE REWRITTEN THIS SECTION TO REFLECT MATHS AND MY PARAMETERIZATION. WE CAN WORK ON SYNCHING OUR NOTATION
%This adds an additional parameter to the system -- the dominance of the drive enhancer in heterozygous males, 
%	when this dominance is high, a narrow sliver of parameter space sustaining a polymorphism emerges (see Figure S\ref{Effect_of_dominance})
 %presumably because positive frequency dependence diminishes with the dominance of the male effect driver.
This male-genotype dependent model requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
When $d_{het}$ is large relative to $d_{hom}$, a narrow sliver of parameter space sustains a polymorphism (see Figure S\ref{Effect_of_dominance}), presumably because positive frequency dependence diminishes with the dominance of the male effect driver. 
As an analytical example, consider the case of male-genotype dependent deleterious recessive female driver 
	(assuming, as above, $w_{BB}=1-s$, and $w_{AA}=w_{AB}=1$). 
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and fixes when $s$ is less than $d_{hom} - 1/2$. 
Intermediate values of $s$, which represent a small proportion of parameter space can maintain a polymorphism.
